---
output: github_document
---

# ggDoE

[![R-CMD-check](https://github.com/toledo60/ggDoE/workflows/R-CMD-check/badge.svg)](https://github.com/toledo60/ggDoE/actions)
[![Documentation](https://img.shields.io/badge/Documentation-ggDoE-blue)](https://ggdoe.netlify.app/)
[![Netlify Status](https://api.netlify.com/api/v1/badges/20d30180-f503-4b63-ba9c-c95bfca3826e/deploy-status)](https://app.netlify.com/sites/ggdoe/deploys)

## Installation

You can get the development version from GitHub:

```{r,eval=FALSE}
#install.packages("remotes") 
remotes::install_github("toledo60/ggDoE")
```

## Overview

With ggDoE you'll be able to generate common plots used in Design of Experiments with ggplot2.

The following plots are currently available:

- [alias_matrix()](https://ggdoe.netlify.app/reference/alias_matrix.html): Correlation matrix plot to visualize the Alias matrix

- [boxcox_transform()](https://ggdoe.netlify.app/reference/boxcox_transform.html): Boxcox transformation plot

- [diagnostic_plots()](https://ggdoe.netlify.app/reference/diagnostic_plots.html): Regression diagnostics plots

- [half_normal()](https://ggdoe.netlify.app/reference/half_normal.html): Half-Normal effects plot


- [interaction_effects()](https://ggdoe.netlify.app/reference/interaction_effects.html): Interaction effects plot between two factors in a factorial design


- [main_effects()](https://ggdoe.netlify.app/reference/main_effects.html): Main effect plots for one factor in a factorial design


- [pareto_plot()](https://ggdoe.netlify.app/reference/pareto_plot.html): Pareto plot of effects with cutoff values for the margin of error (ME) and simultaneous margin of error (SME)



```{r}
library(ggDoE)
```


**Alias Matrix**

```{r}
alias_matrix(design=aliased_design)
```


```{r}
alias_matrix(design=aliased_design, symmetric=TRUE)
```



**Box-Cox Transformation**

```{r}
model <- lm(s2 ~ (A+B+C+D),data = adapted_epitaxial)
boxcox_transform(model,lambda = seq(-2,3,0.2))
```




**Regression Diagnostic Plots**

```{r}
data(mtcars)
mtcars_lm <- lm(mpg ~.,data=mtcars)
diagnostic_plots(mtcars_lm)
```




**Half-Normal Plot**


```{r}
m1 <- lm(lns2 ~ (A+B+C+D)^4,data=original_epitaxial)
half_normal(m1)
```

```{r}
half_normal(m1,method='Zahn',alpha=0.1,
            ref_line=TRUE,label_active=TRUE,
            showME=TRUE)
```


**Interaction Effects Plot (Factorial Design)**


```{r,fig.width=8,fig.height=6}
interaction_effects(adapted_epitaxial,response = 'ybar',
                    exclude_vars = c('s2','lns2'))
```


```{r,fig.width=9,fig.height=4}
interaction_effects(adapted_epitaxial,response = 'ybar',
                    exclude_vars = c('A','s2','lns2'),
                    ncols=3)
```



**Main Effects Plots (Factorial Design)**

```{r}
main_effects(original_epitaxial,
             response='s2',
             exclude_vars = c('ybar','lns2'))
```


```{r}
main_effects(original_epitaxial,
             response='s2',
             exclude_vars = c('A','ybar','lns2'),
             ncols=3)
```



**Pareto Plot**

```{r}
m1 <- lm(lns2 ~ (A+B+C+D)^4,data=original_epitaxial)
pareto_plot(m1)
```


```{r}
pareto_plot(m1,method='Zahn',alpha=0.1)
```



### Contributing to the package
I welcome feedback, suggestions, issues, and contributions!
Check out the [CONTRIBUTING](https://github.com/toledo60/ggDoE/blob/main/.github/CONTRIBUTING.md) file for more details.
